Online {\IterComp}

### 2017-09-15 prof.tex
some of the changes to the structure of the Relaxation Section

In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling the work
counter increments needs to take other basic blocks into account. Figure~\ref{fig:cfg-example} shows an example CFG, where the instrumented
blocks are marked with red. In this example, executing block $B_5$ implies not executing $B_6$. To compute the amount of work associated
with each instrumentation probe, we first need to determine how reaching a probe relates with executing or not a basic block. Starting from
instrumented edges, we build symbolic expressions for each edge of their frequency counts as a linear function of probe counts, as seen in
Figure~\ref{fig:cfg-example}. If a symbolic expression for an edge and the corresponding block contains a positive term for a probe, this
means that the block is executed when we reach the probe, so the block's cost should be added at that probe. Conversely, if the expression
contains a negative term for a probe, then the block is executed when we do not reach the probe, so the block's cost should be subtracted.
In our example CFG, the symbolic expression for the basic block $B_8$ is $P_0 - P_1$, so the amount of work of $B_8$, denoted by $w(B_8)$,
is incremented in probe $P_0$ and decremented in $P_1$.

@@

Optimal instrumentation significantly reduces the profiling overhead when compared to naive instrumentation, from an average overhead of
79\% to 13\%. In some extreme cases though optimal instrumentation can result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d}
in Figure~\ref{fig:overhead-O3}). To further reduce the overhead in these cases, we propose a relaxation strategy that offers a trade-off
between accuracy and overhead. While in traditional basic block profiling the aim is to accurately measure the frequency counts of every
single block, in our case we only care about the total amount of work. Our insight was that if the work associated with a probe is small,
then ignoring the probe will have little effect on the total work estimation. By removing such probes, especially if they are frequently
executed, we can reduce the instrumentation overhead dramatically.

Our relaxation strategy operates on selection of probes and weights as decided by the optimal algorithm. Removing low weight probes is
easy, but we need a way to guarantee that the error in our work estimation will not grow too big. Our conservative approach, decides on an
upper bound for the error and applies it recursively on every loop of the CFG. Specifically, our algorithm extracts all subgraphs
representing loops and transforms them into DAGs (Directed Acyclic Graphs) by removing the backedge and moving inner loops into separate
DAGs. An example of this is shown in Figure~\ref{fig:cfg-relax-example}, where the CFG is partitioned into two DAGs.

@@

\subsection{Whole Program Relaxation}

The proposed relaxation strategy provides hard guarantees that error will remain below the chosen threshold but in many cases these
guarantees are too conservative. Our worst case scenario approach assumes that each loop will always perform the minimum amount of work
possible which might be orders of magnitude less than the work actually performed. In such cases, we remove too few probes and have little
effect on the instrumentation overhead, while the actual relative error in our work estimation is negligible. To avoid this problem, we
propose an adapted version of the relaxation algorithm that operates on the whole program.

\textit{Whole program relaxation} works by using block-frequency profiling from previous executions. With this information, we are able to
compute the error of removing a given probe in terms of the whole program's execution, and then select a subset of all the probes to be
removed.

For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the whole program relaxation as the following 0-1 Knapsack
problem:
\begin{gather*}
\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
x_i\in\{0,1\}, i\in\{0,\ldots,k\}
\end{gather*}
where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $x_i$ denotes the probes selected for removal, $M$ is the
error threshold, and $\varepsilon(P_i)$ is the error of removing probe $P_i$ relative to the total work.
%if $\Delta W$ is the work
%value for the whole program's execution, computed from the basic block frequencies profiled from previous executions, the error for a
%given probe $P_i$ is
%\[
%\varepsilon(P_i) = \frac{\omega(P_i)f(P_i)}{\Delta W}.
%\]

Contrary to the per DAG relaxation, the error introduced by whole program relaxation is not guaranteed to be bounded by the error threshold
$M$. Still, we expect it to remain close to the threshold as long as the program's behavior does not deviate significantly from the
behavior captured in profile we used.

>>>>

In contrast to the na\"ive instrumentation where each basic block records only its own amount of work, with the optimal profiling the work
counter increments needs to take other basic blocks into account. Figure~\ref{fig:cfg-example} shows an example CFG, where the instrumented
blocks are marked with red. In this example, executing block $B_5$ implies not executing $B_6$. To compute the amount of work associated
with each instrumentation probe, we first need to determine how reaching a probe relates with executing or not a basic block. Starting from
instrumented edges, we build symbolic expressions for each edge of their frequency counts as a linear function of probe counts, as seen in
Figure~\ref{fig:cfg-example}. If a symbolic expression for an edge and the corresponding block contains a positive term for a probe, this
means that the block is executed when we reach the probe, so the block's cost should be added at that probe. Conversely, if the expression
contains a negative term for a probe, then the block is executed when we do not reach the probe, so the block's cost should be subtracted.
In our example CFG, the symbolic expression for the basic block $B_8$ is $P_0 - P_1$, so the amount of work of $B_8$, denoted by $w(B_8)$,
is incremented in probe $P_0$ and decremented in $P_1$.

@@

Optimal instrumentation significantly reduces the profiling overhead when compared to na\"ive instrumentation, from an average overhead of
79\% to 13\%. In some extreme cases, optimal instrumentation can result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d}
in Figure~\ref{fig:overhead-O3}). To further reduce the overhead in these cases, we propose two relaxation strategies that offer a trade-off
between accuracy and overhead. Our insight was that if the work associated with a probe is small,
then ignoring the probe will have little effect on the total work estimation.
By removing such probes, especially if they are frequently
executed, we can reduce the instrumentation overhead dramatically.
To contrast with optimal profiling alone,
while the optimal placement of probes tries to reduce overhead by placing probes in edges that are less likely to be executed,
the relaxation strategies focus on removing probes that are more likely to be executed but add little to the work metric.

Our relaxation strategy operates on selection of probes and weights as decided by the optimal algorithm. Removing low weight probes is
easy, but we need a way to guarantee that the error in our work estimation will not grow too big.
To this end, we first propose a more conservative approach that
calculates induced error from removing a probe by considering
    the worst case of any execution path through the program,
and then we propose a more aggressive approach that
    computes induced error based on the relative change of the work measurement that removing a probe would make to the total amount work for the whole program.
We discuss each one of them in detail.

\subsubsection{\WCRelaxTitle Relaxation}

Our conservative approach, decides on an upper bound for the error and applies it recursively on every loop of the CFG. Specifically, our algorithm extracts all subgraphs
representing loops and transforms them into DAGs (Directed Acyclic Graphs) by removing the backedge and moving inner loops into separate
DAGs. An example of this is shown in Figure~\ref{fig:cfg-relax-example}, where the CFG is partitioned into two DAGs.

@@

\subsubsection{\WPRelaxTitle Relaxation}

The \WCRelaxLower relaxation provides hard guarantees that error will remain below the chosen threshold but in many cases these
guarantees are too conservative. It assumes that each loop will always execute the path with the minimum amount of work
possible, which might be orders of magnitude less than the work actually performed. In such cases, we remove too few probes and have little
effect on the profiling overhead, while the actual relative error in our work estimation is negligible. To avoid this problem, we
propose a more aggressive approach that operates on the whole program.

The \WPRelaxLower relaxation works by using block-frequency profiling from previous executions. With this information, we are able to
compute the error of removing a given probe in terms of the whole program's execution, and then select a subset of all the probes to be
removed.

For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, similar to the \WCRelaxLower relaxation,
we model the \WPRelaxLower relaxation as the following 0-1 Knapsack problem:
\begin{gather*}
\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
x_i\in\{0,1\}, i\in\{0,\ldots,k\}
\end{gather*}
where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $x_i$ denotes the probes selected for removal, $M$ is the
error threshold, and $\varepsilon(P_i)$ is the error of removing probe $P_i$ relative to the total work.

Contrary to the per DAG approach, the error introduced by \WPRelaxLower relaxation is not guaranteed to be bounded by the threshold $M$.
Still, we expect it to remain close to the threshold as long as the program's behavior does not deviate significantly from the behavior
captured in profile we used.

### 2017-09-15 related.tex
some of the changes to the structure of the Relaxation Section

Donald Knuth~\cite{knuth73} introduced an optimal algorithm for profiling block frequency, inserting much fewer probes than naive
instrumentation approaches that were used in practice~\cite{knuth71}. Forman~\cite{forman81} proposed further overhead reductions by
placing the probes in basic blocks that are less likely to be executed based on static heuristics. Ball and Larus~\cite{ball94} offered a
detailed discussion comparing two approaches for optimally profiling block frequency, namely, placing probes on the vertices or the edges
of the control-flow graph (CFG). They show that the edge-based approach produces optimal placement of probes. Theirs is the algorithm we
used for our initial selection of probes before relaxation.

>>>>

Donald Knuth~\cite{knuth73} introduced an optimal algorithm for profiling block frequency, inserting much fewer probes than na\"ive
instrumentation approaches that were used in practice~\cite{knuth71}. Forman~\cite{forman81} proposed further overhead reductions by
placing the probes in basic blocks that are less likely to be executed based on static heuristics. Ball and Larus~\cite{ball94} offered a
detailed discussion comparing two approaches for optimally profiling block frequency, namely, placing probes on the vertices or the edges
of the control-flow graph (CFG). They show that the edge-based approach produces optimal placement of probes. Theirs is the algorithm we
used for our initial selection of probes before relaxation.

### 2017-09-15 results.tex
overall changes
some of the changes to the structure of the Relaxation Section

\section{Experimental Results}

@@

@@

The most critical case, amongst the evaluated benchmarks, happens with the
\texttt{adpcm\_d} program, with an overhead of about 59\% for the optimal
instrumentation.
With a threshold of only 5\%, the relaxation reduces
the overhead by a factor of $4.5\times$ over the optimal instrumentation,
and the WPO relaxation by a factor of $4.8\times$.
This benchmark has a single hot function
which consists mainly of a single hot loop with several branches inside it. %, as depicted by Figure~\ref{fig:adpcm_d-cfg-instr}.
Most of the improvement achieved by both strategies comes from removing only two probes from the hot loop. The two relaxed probes were
placed in branches with a high probability of being taken (above 55\% probability), but with a small contribution to the total amount of
work measured in DAG for this loop (both probes with an static error of about 1.3\%). Because these probes were placed in basic blocks with
a considerable execution frequency, removing the probes resulted in a significant reduction in runtime overhead.

@@

\subsubsection{\Itercomp Strategies}
In this experiment, we use evaluate our \itercomp approach against an oracle. We consider five \itercomp strategies:
%When selecting the best optimization sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
\begin{itemize}[leftmargin=3mm]
\item \textbf{Oracle-RM} executes the program twice for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for guiding online {\itercomp}. This oracle provides a reference results for assessing the effectiveness work-efficiency metric.
\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
  It uses the work-efficiency metric for guiding online {\itercomp}.
  Because this version simulates a \textit{perfect} profiling, with zero overhead,
  it allows us isolate and evaluate the effectiveness work-efficiency metric.
\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real-world online scenario.
  The program is executed only once with each input.
  It uses the optimal work profiling for computing the work efficiency.
\item \textbf{Real-R5} is similar to the {Real-OP}, but with a relaxation of 5\% in the work profiling.
\item \textbf{Real-WPO-R5} is similar to the {Real-OP}, but with a whole-program relaxation of 5\% in the work profiling.
%\item \textbf{Real-IPC} uses IPC for guiding the online {\itercomp}.
\end{itemize}

>>>>

\section{Experimental Results}\label{sec:results}

@@

Profiling strategies used throughout this section 
for measuring either the actual speedup or the work efficiency metric.
\begin{itemize}[leftmargin=3mm]
\item \textbf{\OracleRM} measures the actual speedup over the unoptimized version of the program for any given input.
In our experiments, this oracle is allowed to execute each input at least twice so the speedup can be computed.
\item \textbf{\OraclePP} measures the work efficiency metric with a \textit{magically perfect} non-intrusive profiling.
  Because this version simulates a \textit{perfect} profiling, with zero overhead,
  it allows us to isolate and evaluate the actual effectiveness of the work efficiency metric.
\item \textbf{\OptProf} corresponds to the work profiling using the optimal placement of the probes.
  This profiling strategy requires a single execution of the program for any given input.
\item \textbf{\WCRelax-\textit{N}\%} measures the work efficiency metric using the work profiling with a \textit{N}\% thresheld for
the \WCRelaxLower relaxation.
  Similar to the \OptProf, this profiling strategy also requires a single execution of the program for any given input.
\item \textbf{\WPRelax-\textit{N}\%} is similar to the \WCRelax, but it applies the \WPRelaxLower relaxation instead.
\end{itemize}

@@

The most critical case, amongst the evaluated benchmarks, happens with the
\texttt{adpcm\_d} program, with an overhead of about 59\% for the optimal
instrumentation.
With a threshold of only 5\%, the relaxation reduces
the overhead by a factor of $4.5\times$ over the optimal instrumentation,
and the WPO relaxation by a factor of $4.8\times$.
This benchmark has a single hot function
which consists mainly of a single hot loop with several branches inside it. %, as depicted by Figure~\ref{fig:adpcm_d-cfg-instr}.
Most of the improvement achieved by both strategies comes from removing only two probes from the hot loop. The two relaxed probes were
placed in branches with a high probability of being taken, but with a small contribution to the total amount of
work measured in DAG for this loop (both probes with an static error of about 1.3\%). Because these probes were placed in basic blocks with
a considerable execution frequency, removing the probes resulted in a significant reduction in runtime overhead.

@@

\subsubsection{\Itercomp Strategies}
In this experiment, we use evaluate our \itercomp approach against an oracle. We consider five \itercomp strategies:
%When selecting the best optimization sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
\begin{itemize}[leftmargin=3mm]
\item \textbf{Oracle-RM} executes the program twice for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for guiding online {\itercomp}. This oracle provides a reference results for assessing the effectiveness work-efficiency metric.
\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
  It uses the work-efficiency metric for guiding online {\itercomp}.
  Because this version simulates a \textit{perfect} profiling, with zero overhead,
  it allows us isolate and evaluate the effectiveness work-efficiency metric.
\item \textbf{Opt-Prof} corresponds to the online {\itercomp} as it would be applied in a real-world online scenario.
  The program is executed only once with each input.
  It uses the optimal work profiling for computing the work efficiency.
\item \textbf{Real-R5} is similar to the {Real-OP}, but with a relaxation of 5\% in the work profiling.
\item \textbf{Real-WPO-R5} is similar to the {Real-OP}, but with a whole-program relaxation of 5\% in the work profiling.
%\item \textbf{Real-IPC} uses IPC for guiding the online {\itercomp}.
\end{itemize}
